---
title: "BREAST CANCER SURVIVAL PREDICTION FOR DECISION SUPPORT SYSTEM"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}

#install.packages(c("survival", "survminer", "caret", "randomForest", "e1071"))

library(tidyverse)
library(dplyr)
library(survival)
library(survminer)
library(caret)
library(randomForest)
library(e1071)
library(survival)
library(survminer)

```

# DATA CLEANING

```{r libary, message=FALSE, warning=FALSE}
model_data <- read.csv("model_data.csv")
head(data)

```


```{r message=FALSE, warning=FALSE}
colSums(is.na(model_data))
```





## FEATURE ENGINEERING 
```{r preprocessing}
library(dplyr)

# Feature Engineering
fe_model_data <- model_data

# 1. Binning
fe_model_data <- fe_model_data %>%
  mutate(
    Age_Group = cut(Age_at_Diagnosis,
                    breaks = c(0, 30, 45, 60, 75, 100),
                    labels = c("Below_30", "30_45", "45_60", "60_75", "Above_75"),
                    right = FALSE),
    
    Tumor_Size_Group = cut(Tumor_Size,
                    breaks = c(0, 10, 20, 50, 100, 500),
                    labels = c("Tiny", "Small", "Medium", "Large", "Huge"),
                    right = FALSE)
  )

# 2. Interaction Terms
fe_model_data <- fe_model_data %>%
  mutate(
    ER_PR = as.numeric(ER) * as.numeric(PR),
  )

# 3. Reduce rare levels
cancer_type_counts <- table(fe_model_data$Cancer_Type_Detailed)
rare_types <- names(cancer_type_counts[cancer_type_counts < 50])

fe_model_data$Cancer_Type_Detailed <- as.character(fe_model_data$Cancer_Type_Detailed)
fe_model_data$Cancer_Type_Detailed[fe_model_data$Cancer_Type_Detailed %in% rare_types] <- "Other"
fe_model_data$Cancer_Type_Detailed <- as.factor(fe_model_data$Cancer_Type_Detailed)
fe_model_data <- fe_model_data[-1]
head(fe_model_data)
```


# DATA MODELLING FOR SURVIVAL ANALYSIS

```{r}
# Update fe_model_data not model_data!
fe_model_data$Tumor_Stage <- factor(fe_model_data$Tumor_Stage, 
                                    levels = c("Stage I", "Stage II", "Stage III", "Stage IV"))
fe_model_data$ER <- factor(fe_model_data$ER, levels = c(0, 1))
fe_model_data$PR <- factor(fe_model_data$PR, levels = c(0, 1))
fe_model_data$HER2 <- factor(fe_model_data$HER2, levels = c(0, 1))
fe_model_data$Chemo <- factor(fe_model_data$Chemo, levels = c(0, 1))
fe_model_data$Hormone <- factor(fe_model_data$Hormone, levels = c(0, 1))
fe_model_data$Radio <- factor(fe_model_data$Radio, levels = c(0, 1))
fe_model_data$Menopause <- factor(fe_model_data$Menopause, levels = c(0, 1))
# fe_model_data$Age_Group <- factor(fe_model_data$Age_Group, levels = c(1:5), labels = c("Below_30", "30_45", "45_60", "60_75", "Above_75"))
# fe_model_data$Tumor_Size_Group <- factor(fe_model_data$Tumor_Size_Group, levels = c(1:5), labels = c("Tiny", "Small", "Medium", "Large", "Huge"))
fe_model_data$Cancer_Type_Detailed <- factor(fe_model_data$Cancer_Type_Detailed,
                                               levels = c("Breast Invasive Ductal Carcinoma", 
                                                          "Breast Invasive Lobular Carcinoma",
                                                          "Breast Mixed Ductal and Lobular Carcinoma", "Other"))

fe_model_data$ER_PR <- factor(fe_model_data$ER_PR, levels = c(0, 1))



# Ensure overall survival is numeric
fe_model_data$Overall_Survival_Months_ <- as.numeric(fe_model_data$Overall_Survival_Months_)
```


## DATA PREPROCESSING AND SPLITTING
```{r message=FALSE, warning=FALSE}
library(caret)

# Prepare binary classification targets
fe_model_data$Surv_Status <- factor(fe_model_data$Surv_Status, levels = c(0, 1))
fe_model_data$Relapse_Status <- factor(fe_model_data$Relapse_Status, levels = c(0, 1))

# Split into train/test
set.seed(123)
train_index <- createDataPartition(fe_model_data$Surv_Status, p = 0.8, list = FALSE)

train_data <- fe_model_data[train_index, ]
test_data <- fe_model_data[-train_index, ]

# Set targets
ss_train <- train_data$Surv_Status
rs_train <- train_data$Relapse_Status
x_train <- train_data[, setdiff(names(train_data), c("Surv_Status", "Overall_Survival_Months_", "Relapse_Status", "Relapse_Months"))]

ss_test <- test_data$Surv_Status
rs_test <- test_data$Relapse_Status
x_test <- test_data[, setdiff(names(test_data), c("Surv_Status", "Overall_Survival_Months_", "Relapse_Status", "Relapse_Months"))]

# For survival months prediction
sm_train <- train_data$Overall_Survival_Months_
# sm_test <- test_data$Overall_Survival_Months_
rm_train <- train_data$Relapse_Months
# rm_test <- test_data$Relapse_Months

x_train_m <- x_train
x_test_m <- x_test

```







# TRAINING MODELS

## OVERALL SURVIVAL STATUS

```{r lr}
ss_log_reg_model <- glm(ss_train ~ .,
                     data = x_train, family = "binomial")

# Summary of the model
summary(ss_log_reg_model)


# Predict on test data
ss_predictions_status <- predict(ss_log_reg_model, newdata = x_test, type = "response")
ss_pred_class_status <- ifelse(ss_predictions_status > 0.5, 1, 0)  # 1 = Dead, 0 = Alive

# Confusion Matrix to evaluate the performance
ss_conf_matrix_status <- confusionMatrix(factor(ss_pred_class_status), factor(ss_test))
print(ss_conf_matrix_status)

```


### Random Forest
```{r RF}
library(randomForest)

set.seed(123)
ss_rf_model <- randomForest(ss_train ~ .,
                         data = x_train, ntree = 500, importance = TRUE)


# Predict and evaluate
ss_rf_prop <- predict(ss_rf_model, newdata = x_test, type = "prob")
# Pick the class (column name) with the highest probability for each observation
ss_rf_pred <- colnames(ss_rf_prop)[max.col(ss_rf_prop, ties.method = "first")]
ss_rf_pred <- factor(ss_rf_pred, levels = c("0", "1"))



confusionMatrix(ss_rf_pred, factor(ss_test))

```



```{r glm}
library(glmnet)

# Create matrices
x_train_matrix <- model.matrix(~ . -1, data = x_train)
x_test_matrix <- model.matrix(~ . -1, data = x_test)

# Fit model (alpha=0.5 for Elastic Net)
ss_cv_model <- cv.glmnet(x_train_matrix, ss_train, alpha = 0.5, family = "binomial")

# Predict
ss_enet_pred <- predict(ss_cv_model, newx = x_test_matrix, s = "lambda.min", type = "response")
ss_enet_pred_class <- ifelse(enet_pred > 0.5, 1, 0)
confusionMatrix(as.factor(ss_enet_pred_class), as.factor(ss_test))
```



```{r xgboost}

library(xgboost)


ss_xgb_model <- xgboost(data = x_train_matrix, label = as.numeric(ss_train) - 1,
                     objective = "binary:logistic", nrounds = 100, max.depth = 6, eta = 0.1)

# Prediction
ss_xgb_pred <- predict(ss_xgb_model, newdata = x_test_matrix)
ss_xgb_pred_class <- ifelse(xgb_pred > 0.5, 1, 0)
confusionMatrix(as.factor(ss_xgb_pred_class), ss_test)


```









## OVERALL SURVIVAL  MONTHS
```{r}

# Train a Linear Regression model for Overall Survival Months
sm_linear_reg_model <- lm(sm_train ~ .,
                       data = x_train_m)

# Summary of the model
summary(sm_linear_reg_model)

# Predict on test data
sm_predictions_osm <- predict(sm_linear_reg_model, newdata = x_test_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
sm_rmse_osm <- sqrt(mean((sm_predictions_osm - sm_test)^2))
print(paste("RMSE for Linear Regression:", sm_rmse_osm))

```



```{r}
library(randomForest)

# Train a Random Forest Regression model
sm_rf_reg_model <- randomForest(sm_train ~ .,
                             data = x_train_m, ntree = 500)

# Predict on test data
sm_rf_reg_pred <- predict(sm_rf_reg_model, newdata = x_test_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
sm_rf_reg_rmse <- sqrt(mean((sm_rf_reg_pred - sm_test)^2))
print(paste("RMSE for Random Forest Regression:", sm_rf_reg_rmse))
```

```{r glmt}
library(glmnet)

# Create matrices
x_train_matrix_m <- model.matrix(~ . -1, data = x_train_m)
x_test_matrix_m <- model.matrix(~ . -1, data = x_test_m)

# Create matrices
# Fit model (alpha=0.5 for Elastic Net)
sm_cv_model_osm <- cv.glmnet(x_train_matrix_m, sm_train, alpha = 0.5)

# Predict
sm_enet_pred_osm <- predict(sm_cv_model_osm, newx = x_test_matrix_m, s = "lambda.min")

# Evaluate performance with RMSE (Root Mean Squared Error)
sm_enet_rmse_osm <- sqrt(mean((sm_enet_pred_osm - sm_test)^2))
print(paste("RMSE for Elastic Net Regression:", sm_enet_rmse_osm))
```



```{r xgboost}
library(xgboost)

# Train an XGBoost model
sm_xgb_model_osm <- xgboost(data = x_train_matrix_m, label = sm_train,
                         objective = "reg:squarederror", nrounds = 100, max.depth = 6, eta = 0.1)

# Predict
sm_xgb_pred_osm <- predict(sm_xgb_model_osm, newdata = x_test_matrix_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
sm_xgb_rmse_osm <- sqrt(mean((sm_xgb_pred_osm - sm_test)^2))
print(paste("RMSE for XGBoost Regression:", sm_xgb_rmse_osm))
```





## RELAPSE STATUS MODEL TRAINING

```{r lr}
rs_log_reg_model <- glm(rs_train ~ .,
                     data = x_train, family = "binomial")

# Summary of the model
summary(rs_log_reg_model)


# Predict on test data
predictions_status <- predict(rs_log_reg_model, newdata = x_test, type = "response")
pred_class_status <- ifelse(predictions_status > 0.5, 1, 0)  # 1 = Dead, 0 = Alive

# Confusion Matrix to evaluate the performance
rs_conf_matrix_status <- confusionMatrix(factor(pred_class_status), factor(rs_test))
print(rs_conf_matrix_status)

```


### Random Forest
```{r RF}
library(randomForest)

set.seed(123)
rs_rf_model <- randomForest(rs_train ~ .,
                         data = x_train, ntree = 500, importance = TRUE)

# Predict and evaluate
rs_rf_pred <- predict(rs_rf_model, newdata = x_test)

confusionMatrix(rs_rf_pred, factor(rs_test))

```



```{r glm}
library(glmnet)

# Fit model (alpha=0.5 for Elastic Net)
rs_cv_model <- cv.glmnet(x_train_matrix, rs_train, alpha = 0.5, family = "binomial")

# Predict
enet_pred <- predict(rs_cv_model, newx = x_test_matrix, s = "lambda.min", type = "response")
enet_pred_class <- ifelse(enet_pred > 0.5, 1, 0)
confusionMatrix(as.factor(enet_pred_class), as.factor(rs_test))
```



```{r xgboost}

library(xgboost)


rs_xgb_model <- xgboost(data = x_train_matrix, label = as.numeric(rs_train) - 1,
                     objective = "binary:logistic", nrounds = 100, max.depth = 6, eta = 0.1)

# Prediction
xgb_pred <- predict(rs_xgb_model, newdata = x_test_matrix)
xgb_pred_class <- ifelse(xgb_pred > 0.5, 1, 0)
confusionMatrix(as.factor(xgb_pred_class), factor(rs_test))

```







## OVERALL RELAPSE  MONTHS
```{r lr}
rm_linear_reg_model <- lm(rm_train ~ .,
                       data = x_train_m)

# Summary of the model
summary(rm_linear_reg_model)

# Predict on test data
rm_predictions_osm <- predict(rm_linear_reg_model, newdata = x_train_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
rm_rmse_osm <- sqrt(mean((rm_predictions_osm - rm_train)^2))
print(paste("RMSE for Linear Regression:", rm_rmse_osm))

```



```{r}
library(randomForest)

# Train a Random Forest Regression model
rm_rf_reg_model <- randomForest(rm_train ~ .,
                             data = x_train_m, ntree = 500)

# Predict on test data
rm_rf_reg_pred <- predict(rm_rf_reg_model, newdata = x_train_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
rm_rf_reg_rmse <- sqrt(mean((rm_rf_reg_pred - rm_train)^2))
print(paste("RMSE for Random Forest Regression:", rm_rf_reg_rmse))
```

```{r}
library(glmnet)

# Create matrices
# Fit model (alpha=0.5 for Elastic Net)
rm_cv_model_osm <- cv.glmnet(x_train_matrix_m, rm_train, alpha = 0.5)

# Predict
rm_enet_pred_osm <- predict(cv_model_osm, newx = x_train_matrix_m, s = "lambda.min")

# Evaluate performance with RMSE (Root Mean Squared Error)
rm_enet_rmse_osm <- sqrt(mean((rm_enet_pred_osm - rm_train)^2))
print(paste("RMSE for Elastic Net Regression:", rm_enet_rmse_osm))
```



```{r xgboost}
library(xgboost)

# Train an XGBoost model
rm_xgb_model_osm <- xgboost(data = x_train_matrix_m, label = rm_train,
                         objective = "reg:squarederror", nrounds = 100, max.depth = 6, eta = 0.1)

# Predict
rm_xgb_pred_osm <- predict(rm_xgb_model_osm, newdata = x_train_matrix_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
rm_xgb_rmse_osm <- sqrt(mean((rm_xgb_pred_osm - rm_test)^2))
print(paste("RMSE for XGBoost Regression:", rm_xgb_rmse_osm))
```







```{r classification_metric}
# Collect classification results
classification_results <- data.frame(
  Model = c(
    "Overall Survival - Logistic Regression",
    "Overall Survival - Random Forest",
    "Overall Survival - Elastic Net",
    "Overall Survival - XGBoost",
    "Relapse Status - Logistic Regression",
    "Relapse Status - Random Forest",
    "Relapse Status - Elastic Net",
    "Relapse Status - XGBoost"
  ),
  Accuracy = c(
    conf_matrix_status$overall["Accuracy"],
    confusionMatrix(ss_rf_pred, factor(ss_test))$overall["Accuracy"],
    confusionMatrix(as.factor(ss_enet_pred_class), as.factor(ss_test))$overall["Accuracy"],
    confusionMatrix(as.factor(ss_xgb_pred_class), ss_test)$overall["Accuracy"],
    rs_conf_matrix_status$overall["Accuracy"],
    confusionMatrix(rs_rf_pred, factor(rs_test))$overall["Accuracy"],
    confusionMatrix(as.factor(enet_pred_class), as.factor(rs_test))$overall["Accuracy"],
    confusionMatrix(as.factor(xgb_pred_class), factor(rs_test))$overall["Accuracy"]
  ),
  Kappa = c(
    conf_matrix_status$overall["Kappa"],
    confusionMatrix(rf_pred, factor(ss_test))$overall["Kappa"],
    confusionMatrix(as.factor(enet_pred_class), as.factor(ss_test))$overall["Kappa"],
    confusionMatrix(as.factor(xgb_pred_class), ss_test)$overall["Kappa"],
    rs_conf_matrix_status$overall["Kappa"],
    confusionMatrix(rs_rf_pred, factor(rs_test))$overall["Kappa"],
    confusionMatrix(as.factor(enet_pred_class), as.factor(rs_test))$overall["Kappa"],
    confusionMatrix(as.factor(xgb_pred_class), factor(rs_test))$overall["Kappa"]
  )
)

# Print the table
print(classification_results)

```


```{r}
# Collect regression results
regression_results <- data.frame(
  Model = c(
    "Overall Survival Months - Linear Regression",
    "Overall Survival Months - Random Forest",
    "Overall Survival Months - Elastic Net",
    "Overall Survival Months - XGBoost",
    "Relapse Months - Linear Regression",
    "Relapse Months - Random Forest",
    "Relapse Months - Elastic Net",
    "Relapse Months - XGBoost"
  ),
  RMSE = c(
    sm_rmse_osm,
    sm_rf_reg_rmse,
    sm_enet_rmse_osm,
    sm_xgb_rmse_osm,
    rm_rmse_osm,
    rm_rf_reg_rmse,
    rm_enet_rmse_osm,
    rm_xgb_rmse_osm
  )
)

# Print the table
print(regression_results)

```



## SAVING THE BEST MODEL
```{r}
saveRDS(ss_rf_model, "ss_rf_model.rds")
saveRDS(rf_reg_model, "sm_rf_reg_model.rds")
saveRDS(rs_cv_model, "rs_cv_model.rds")
saveRDS(rs_xgb_model, "rs_xgb_model.rds")
saveRDS(rm_rf_reg_model, "rm_rf_reg_model.rds")
```




# SIMULATION

```{r new_data}
# Simulate a new data point for prediction
new_data <- data.frame(
  Age_at_Diagnosis = 50,
  Tumor_Size = 3.5,  # Example size in cm
  Cancer_Type_Detailed = 3,  # E.g., "Breast Invasive Ductal Carcinoma"
  Lymph_nodes_examined_positive = 2,
  Nottingham_prognostic_index = 4.5,
  Tumor_Stage = 2,  # Tumor stage as a factor
  Neoplasm_Histologic_Grade = 2,  # Example grade
  ER = 1,  # ER positive
  PR = 1,  # PR positive
  HER2 = 0,  # HER2 negative
  Chemo = 1,  # Chemotherapy received
  Hormone = 0,  # Hormone therapy not received
  Radio = 1,  # Radiotherapy received
  Menopause = 1  # Post-menopausal
)



# View the new data
new_data
```


```{r}

# Convert Tumor Stage to a factor with appropriate labels
new_data$Tumor_Stage <- factor(new_data$Tumor_Stage,
                                levels = c(1, 2, 3, 4),
                                labels = c("Stage I", "Stage II", "Stage III", "Stage IV"))


# Encode ER/PR/HER2 status as binary (Positive = 1, Negative = 0)
new_data$ER <- factor(new_data$ER, levels = c(0, 1))
new_data$PR <- factor(new_data$PR,levels = c(0, 1))
new_data$HER2 <- factor(new_data$HER2, levels = c(0, 1))

new_data$Neoplasm_Histologic_Grade <- factor(new_data$Neoplasm_Histologic_Grade, levels = c(1:3))

# Encode Menopausal state as binary
new_data$Menopause <- factor(new_data$Menopause, levels = c(0, 1))

# Encode therapies (Chemotherapy, Hormone Therapy, Radiotherapy) as binary
new_data$Chemo <- factor(new_data$Chemo, levels = c(0, 1))
new_data$Hormone <- factor(new_data$Hormone, levels = c(0, 1))
new_data$Radio <- factor(new_data$Radio, levels = c(0, 1))

# Convert Cancer Type Detailed to factor and numeric
new_data$Cancer_Type_Detailed <- factor(new_data$Cancer_Type_Detailed, levels = c(1:7),
                                         labels = c("Breast", "Breast Invasive Ductal Carcinoma",
                                                    "Breast Invasive Lobular Carcinoma", "Breast Invasive Mixed Mucinous Carcinoma",
                                                    "Breast Mixed Ductal and Lobular Carcinoma", "Invasive Breast Carcinoma", "Others"))




# Ensure necessary columns are present
new_data <- new_data %>%
  select(Age_at_Diagnosis, Tumor_Size, Cancer_Type_Detailed,
         Lymph_nodes_examined_positive, Nottingham_prognostic_index, Tumor_Stage, 
         Neoplasm_Histologic_Grade, ER, PR, HER2, Chemo, Hormone, Radio, Menopause)
head(new_data)
```







### Relapse Free Status
```{r}
# Convert new_data to a numeric matrix
new_x <- model.matrix(~ . - 1, data = new_data)

# Now predict
rs_prediction <- predict(rs_cv_model, newx = new_x, s = "lambda.min", type = "response")

# Output the predicted survival months
sm_prediction
```


### Relapse Free Months
```{r}

# Predict survival status using the pre-trained Elastic Net model (cv_model)
rm_prediction <- predict(rm_rf_reg_model, newdata = new_data)


# Output the predicted survival status
ss_prediction
```








### Survival Months
```{r}

#setequal(colnames(x_new_data), colnames(x_train_matrix))  # Should return TRUE

# Predict survival months using the pre-trained XGBoost model (xgb_model_osm)
sm_prediction <- predict(rf_reg_model, newdata = new_data)

# Output the predicted survival months
sm_prediction
```


### Survival  Status
```{r}
# Create matrix for Elastic Net prediction of Survival Status
x_new_data <- model.matrix(~ . -1, data = new_data)

# Rename new data columns to match training data
colnames(x_new_data)[colnames(x_new_data) == "ERYes"] <- "ER1"
colnames(x_new_data)[colnames(x_new_data) == "PRYes"] <- "PR1"
colnames(x_new_data)[colnames(x_new_data) == "HER2Yes"] <- "HER21"
colnames(x_new_data)[colnames(x_new_data) == "ChemoYes"] <- "Chemo1"
colnames(x_new_data)[colnames(x_new_data) == "HormoneYes"] <- "Hormone1"
colnames(x_new_data)[colnames(x_new_data) == "RadioYes"] <- "Radio1"
colnames(x_new_data)[colnames(x_new_data) == "MenopausePost"] <- "Menopause1"


# Predict survival status using the pre-trained Elastic Net model (cv_model)
ss_prediction <- predict(ss_rf_model, newdata = new_data)


# Output the predicted survival status
ss_prediction
```





