---
title: "BREAST CANCER SURVIVAL PREDICTION FOR DECISION SUPPORT SYSTEM"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}

#install.packages(c("survival", "survminer", "caret", "randomForest", "e1071"))

library(tidyverse)
library(dplyr)
library(survival)
library(survminer)
library(caret)
library(randomForest)
library(e1071)
library(survival)
library(survminer)

```

# DATA CLEANING

```{r libary, message=FALSE, warning=FALSE}
data <- read.csv("Breast Cancer METABRIC.csv")
head(data)

```



```{r message=FALSE, warning=FALSE}
colSums(is.na(data))

```



```{r message=FALSE, warning=FALSE}
colnames(data)
```




```{r message=FALSE, warning=FALSE}
selected_cols <- c(
  "Patient.ID",
  "Age.at.Diagnosis", "Inferred.Menopausal.State",
  "Tumor.Size", "Tumor.Stage", "Neoplasm.Histologic.Grade", 
  "Lymph.nodes.examined.positive", "Cancer.Type.Detailed",
  "Nottingham.prognostic.index",
  "Pam50...Claudin.low.subtype", "ER.Status", "PR.Status", 
  "HER2.Status", "X3.Gene.classifier.subtype",
  "Chemotherapy", "Hormone.Therapy", "Radio.Therapy",
  "Overall.Survival..Months.", "Overall.Survival.Status", 
  "Relapse.Free.Status", "Relapse.Free.Status..Months."
)
# Subset the data
metabric_sub <- data %>%
  select(all_of(selected_cols))

colSums(is.na(metabric_sub))

head(data)
```



### Handling Missing Values
```{r message=FALSE, warning=FALSE}
# Drop only the rows where Overall.Survival.Months_ is NA
# metabric_sub$Age.at.Diagnosis[is.na(metabric_sub$Age.at.Diagnosis)] <- median(metabric_sub$Age.at.Diagnosis, na.rm = TRUE)
# metabric_sub$Tumor.Size[is.na(metabric_sub$Tumor.Size)] <- median(metabric_sub$Tumor.Size, na.rm = TRUE)
# metabric_sub$Tumor.Stage[is.na(metabric_sub$Tumor.Stage)] <- "Unknown"
# mode_grade <- names(sort(table(metabric_sub$Neoplasm.Histologic.Grade), decreasing = TRUE))[1]
# metabric_sub$Neoplasm.Histologic.Grade[is.na(metabric_sub$Neoplasm.Histologic.Grade)] <- mode_grade
# metabric_sub$Lymph.nodes.examined.positive[is.na(metabric_sub$Lymph.nodes.examined.positive)] <- median(metabric_sub$Lymph.nodes.examined.positive, na.rm = TRUE)
# metabric_sub$Nottingham.prognostic.index[is.na(metabric_sub$Nottingham.prognostic.index)] <- median(metabric_sub$Nottingham.prognostic.index, na.rm = TRUE)


mode_value <- "ER+/HER2- Low Prolif"

# Replace blank spaces or NAs in 'X3.Gene.classifier.subtype' with the mode value
metabric_sub$X3.Gene.classifier.subtype[metabric_sub$X3.Gene.classifier.subtype == ""] <- mode_value
  
# Check if replacement was successful
table(metabric_sub$X3.Gene.classifier.subtype)


metabric_sub <- na.omit(metabric_sub)

colSums(is.na(metabric_sub))
dim(metabric_sub)

```


```{r message=FALSE, warning=FALSE}

# Clean column names for easier reference
colnames(metabric_sub) <- gsub("[\\.]+", "_", colnames(metabric_sub))
# View(metabric_sub)
head(metabric_sub)

```


## Encode Categorical Variables with Clinical Coding Where Applicable
We'll encode key variables based on clinical standards, e.g.,:
Tumor Grade (Histologic): Convert to ordinal (1, 2, 3)
Tumor Stage: Standard TNM categories (Stage I, II, III, IV)
ER/PR/HER2 Status: Binary (Positive = 1, Negative = 0)
Treatment flags: Binary (Yes = 1, No = 0)

```{r message=FALSE, warning=FALSE}
library(dplyr)

# Remove rows with missing values
metabric_clean <- na.omit(metabric_sub)

# Binary encoding for Overall Survival
metabric_clean$Surv_Status <- ifelse(metabric_clean$Overall_Survival_Status == "Deceased", 1, 0)

# Binary encoding for Relapse Status (Recurred = 1, Not Recurred = 0)
metabric_clean$Relapse_Status <- ifelse(metabric_clean$Relapse_Free_Status == "Recurred", 1, 0)

# Rename Relapse-Free Survival Months
metabric_clean$Relapse_Months <- metabric_clean$Relapse_Free_Status_Months_

# Encode Tumor Stage as ordered factor
metabric_clean$Tumor_Stage <- factor(metabric_clean$Tumor_Stage,
                                     levels = c(1, 2, 3, 4),
                                     labels = c("Stage I", "Stage II", "Stage III", "Stage IV"))

# Treat 0 stage as missing if applicable
metabric_clean$Tumor_Stage[metabric_clean$Tumor_Stage == "0"] <- NA

# Encode ER/PR/HER2 status
metabric_clean$ER <- ifelse(metabric_clean$ER_Status == "Positive", 1, 0)
metabric_clean$PR <- ifelse(metabric_clean$PR_Status == "Positive", 1, 0)
metabric_clean$HER2 <- ifelse(metabric_clean$HER2_Status == "Positive", 1, 0)

# Encode Menopausal State
metabric_clean$Menopause <- ifelse(metabric_clean$Inferred_Menopausal_State == "Post", 1, 0)

# Encode therapies
metabric_clean$Chemo <- ifelse(metabric_clean$Chemotherapy == "Yes", 1, 0)
metabric_clean$Hormone <- ifelse(metabric_clean$Hormone_Therapy == "Yes", 1, 0)
metabric_clean$Radio <- ifelse(metabric_clean$Radio_Therapy == "Yes", 1, 0)

# Factorize Cancer Type
metabric_clean$Cancer_Type_Detailed <- factor(metabric_clean$Cancer_Type_Detailed, levels = c("Breast", "Breast Invasive Ductal Carcinoma",
                                                    "Breast Invasive Lobular Carcinoma", "Breast Invasive Mixed Mucinous Carcinoma",
                                                    "Breast Mixed Ductal and Lobular Carcinoma", "Invasive Breast Carcinoma", "Others"))

# Optional: Check the mapping
table(metabric_clean$Cancer_Type_Detailed)

# Drop NA rows (after transformations)
metabric_clean <- na.omit(metabric_clean)

# Final model-ready dataset including relapse variables
model_data <- metabric_clean %>%
  select(Surv_Status, Overall_Survival_Months_, Relapse_Status, Relapse_Months,
         Age_at_Diagnosis, Tumor_Size, Cancer_Type_Detailed,
         Lymph_nodes_examined_positive, Nottingham_prognostic_index, Tumor_Stage, 
         Neoplasm_Histologic_Grade, ER, PR, HER2, Chemo, Hormone, Radio, Menopause)

# Preview data
head(model_data)

```




```{r message=FALSE, warning=FALSE}
colSums(is.na(model_data))
write.csv(model_data, 'model_data.csv')
```







# DATA MODELLING FOR SURVIVAL ANALYSIS

```{r}
model_data$Tumor_Stage <- factor(model_data$Tumor_Stage, 
                                 levels = c("Stage I", "Stage II", "Stage III", "Stage IV"))

model_data$ER <- factor(model_data$ER, levels = c(0, 1))  # Estrogen Receptor status
model_data$PR <- factor(model_data$PR, levels = c(0, 1))  # Progesterone Receptor status
model_data$HER2 <- factor(model_data$HER2, levels = c(0, 1))  # HER2 status
model_data$Chemo <- factor(model_data$Chemo, levels = c(0, 1))  # Chemotherapy status
model_data$Hormone <- factor(model_data$Hormone, levels = c(0, 1))  # Hormone therapy status
model_data$Radio <- factor(model_data$Radio, levels = c(0, 1))  # Radiotherapy status
model_data$Menopause <- factor(model_data$Menopause, levels = c(0, 1))  # Menopausal state
# Convert Cancer Type to a factor
model_data$Cancer_Type_Detailed <- factor(model_data$Cancer_Type_Detailed)
model_data$Neoplasm_Histologic_Grade <- factor(model_data$Neoplasm_Histologic_Grade, levels = c(1:3))
model_data$Cancer_Type_Detailed <- factor(model_data$Cancer_Type_Detailed,
                                         levels = c("Breast", "Breast Invasive Ductal Carcinoma",
                                                    "Breast Invasive Lobular Carcinoma", "Breast Invasive Mixed Mucinous Carcinoma",
                                                    "Breast Mixed Ductal and Lobular Carcinoma", "Invasive Breast Carcinoma", "Others"))

# Ensure Overall_Survival_Months_ is numeric
model_data$Overall_Survival_Months_ <- as.numeric(model_data$Overall_Survival_Months_)

head(model_data)
```
```{r EDA}
# Plotting Overall Survival vs Survival Status
ggplot(model_data, aes(x = Surv_Status, y = Overall_Survival_Months_)) +
  geom_boxplot(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Overall Survival Duration by Survival Status",
       x = "Survival Status", y = "Overall Survival Duration (Months)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))



# Plotting Relapse Duration vs Relapse Status
ggplot(model_data, aes(x = Relapse_Status, y = Relapse_Months)) +
  geom_boxplot(fill = "lightcoral", color = "black", alpha = 0.7) +
  labs(title = "Relapse Duration by Relapse Status",
       x = "Relapse Status", y = "Relapse Duration (Months)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))


```
```{r}
# Plotting distribution of Relapse Status with labels
ggplot(model_data, aes(x = Relapse_Status, fill = Relapse_Status)) +
  geom_bar(color = "black", width = 0.7) +  # Bar plot with black outline
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 4) +  # Adding count labels on top of bars
  labs(title = "Distribution of Relapse Status",
       x = "Relapse Status", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),  # Title styling
        axis.title = element_text(size = 12),  # Axis label styling
        axis.text = element_text(size = 10),  # Axis text styling
        legend.position = "none") +  # Hiding legend (optional)
  scale_fill_manual(values = c("0" = "green", "1" = "red"))

```

```{r}

# Plotting distribution of Survival Status with labels
ggplot(model_data, aes(x = Surv_Status, fill = Surv_Status)) +
  geom_bar(color = "black", width = 0.7) +  # Bar plot with black outline
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 4) +  # Adding count labels on top of bars
  labs(title = "Distribution of Survival Status",
       x = "Survival Status", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),  # Title styling
        axis.title = element_text(size = 12),  # Axis label styling
        axis.text = element_text(size = 10),  # Axis text styling
        legend.position = "none") +  # Hiding legend (optional)
  scale_fill_manual(values = c("0" = "blue", "1" = "orange"))
```



## DATA PREPROCESSING AND SPLITTING
```{r message=FALSE, warning=FALSE}
# Prepare the data for binary classification (Survival Status)
model_data$Surv_Status <- factor(model_data$Surv_Status, levels = c(0, 1))  # Alive (0) or Dead (1)
model_data$Relapse_Status <- factor(model_data$Relapse_Status, levels = c(0, 1))  


# Separate predictors and target variable
sm_train <- model_data$Overall_Survival_Months_
rm_train <- model_data$Relapse_Months
x_train_m <- model_data[, setdiff(names(model_data), c("Surv_Status", "Overall_Survival_Months_", "Relapse_Status", "Relapse_Months"))]

sm_test <- model_data$Overall_Survival_Months_
rm_test <- model_data$Relapse_Months
x_test_m <- model_data[, setdiff(names(model_data), c("Surv_Status", "Overall_Survival_Months_", "Relapse_Status", "Relapse_Months"))]



# Split data into training and testing sets (80% training, 20% testing)
set.seed(123)
train_index <- createDataPartition(model_data$Surv_Status, p = 0.8, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

# Convert Surv_Status to factor for classification
train_data$Surv_Status <- as.factor(train_data$Surv_Status)
test_data$Surv_Status <- as.factor(test_data$Surv_Status)

# Separate predictors and target variable
ss_train <- train_data$Surv_Status
#sm_train <- train_data$Overall_Survival_Months_
rs_train <- train_data$Relapse_Status
#rm_train <- train_data$Relapse_Months
x_train <- train_data[, setdiff(names(train_data), c("Surv_Status", "Overall_Survival_Months_", "Relapse_Status", "Relapse_Months"))]

ss_test <- test_data$Surv_Status
#sm_test <- test_data$Overall_Survival_Months_
rs_test <- test_data$Relapse_Status
#rm_test <- test_data$Relapse_Months
x_test <- test_data[, setdiff(names(test_data), c("Surv_Status", "Overall_Survival_Months_", "Relapse_Status", "Relapse_Months"))]

  

```


```{r}
# Calculate the number of samples in training and test datasets
train_size <- nrow(train_data)
test_size <- nrow(test_data)

# Create a data frame for plotting
sample_sizes <- data.frame(
  Dataset = c("Training Data", "Test Data"),
  Sample_Size = c(train_size, test_size)
)

# Plotting the sample sizes
ggplot(sample_sizes, aes(x = Dataset, y = Sample_Size, fill = Dataset)) +
  geom_bar(stat = "identity", color = "black", width = 0.5) +
  geom_text(aes(label = Sample_Size), vjust = -0.5, size = 5) +
  labs(title = "Comparison of Training and Test Sample Sizes",
       x = "Dataset", y = "Sample Size") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10)) +
  scale_fill_manual(values = c("Training Data" = "blue", "Test Data" = "orange"))

```

# TRAINING MODELS

## OVERALL SURVIVAL STATUS

```{r lr}
ss_log_reg_model <- glm(ss_train ~ .,
                     data = x_train, family = "binomial")

# Summary of the model
summary(ss_log_reg_model)


# Predict on test data
ss_predictions_status <- predict(ss_log_reg_model, newdata = x_test, type = "response")
ss_pred_class_status <- ifelse(ss_predictions_status > 0.5, 1, 0)  # 1 = Dead, 0 = Alive

# Confusion Matrix to evaluate the performance
ss_conf_matrix_status <- confusionMatrix(factor(ss_pred_class_status), factor(ss_test))
print(ss_conf_matrix_status)

```


### Random Forest
```{r rf}
library(randomForest)

set.seed(123)
ss_rf_model <- randomForest(ss_train ~ Age_at_Diagnosis + Tumor_Size + Cancer_Type_Detailed + 
                         Lymph_nodes_examined_positive + Nottingham_prognostic_index + 
                         Tumor_Stage + Neoplasm_Histologic_Grade + ER + PR + HER2 + 
                         Chemo + Hormone + Radio + Menopause, 
                         data = x_train, ntree = 500, importance = TRUE)


# Predict and evaluate
ss_rf_prop <- predict(ss_rf_model, newdata = x_test, type = "prob")
# Pick the class (column name) with the highest probability for each observation
ss_rf_pred <- colnames(ss_rf_prop)[max.col(ss_rf_prop, ties.method = "first")]
ss_rf_pred <- factor(ss_rf_pred, levels = c("0", "1"))



confusionMatrix(ss_rf_pred, factor(ss_test))

```



```{r glm}
library(glmnet)

# Create matrices
x_train_matrix <- model.matrix(~ . -1, data = x_train)
x_test_matrix <- model.matrix(~ . -1, data = x_test)

# Fit model (alpha=0.5 for Elastic Net)
ss_cv_model <- cv.glmnet(x_train_matrix, ss_train, alpha = 0.5, family = "binomial")

# Predict
ss_enet_pred <- predict(ss_cv_model, newx = x_test_matrix, s = "lambda.min", type = "response")
ss_enet_pred_class <- ifelse(ss_enet_pred > 0.5, 1, 0)
confusionMatrix(as.factor(ss_enet_pred_class), as.factor(ss_test))
```



```{r xgboost}

library(xgboost)


ss_xgb_model <- xgboost(data = x_train_matrix, label = as.numeric(ss_train) - 1,
                     objective = "binary:logistic", nrounds = 100, max.depth = 6, eta = 0.1)

# Prediction
ss_xgb_pred <- predict(ss_xgb_model, newdata = x_test_matrix)
ss_xgb_pred_class <- ifelse(ss_xgb_pred > 0.5, 1, 0)
confusionMatrix(as.factor(ss_xgb_pred_class), ss_test)


```









## OVERALL SURVIVAL  MONTHS
```{r}

# Train a Linear Regression model for Overall Survival Months
sm_linear_reg_model <- lm(sm_train ~ .,
                       data = x_train_m)

# Summary of the model
summary(sm_linear_reg_model)

# Predict on test data
sm_predictions_osm <- predict(sm_linear_reg_model, newdata = x_test_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
sm_rmse_osm <- sqrt(mean((sm_predictions_osm - sm_test)^2))
print(paste("RMSE for Linear Regression:", sm_rmse_osm))

```



```{r}
library(randomForest)

# Train a Random Forest Regression model
sm_rf_reg_model <- randomForest(sm_train ~ .,
                             data = x_train_m, ntree = 500)

# Predict on test data
sm_rf_reg_pred <- predict(sm_rf_reg_model, newdata = x_test_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
sm_rf_reg_rmse <- sqrt(mean((sm_rf_reg_pred - sm_test)^2))
print(paste("RMSE for Random Forest Regression:", sm_rf_reg_rmse))
```

```{r glmt}
library(glmnet)

# Create matrices
x_train_matrix_m <- model.matrix(~ . -1, data = x_train_m)
x_test_matrix_m <- model.matrix(~ . -1, data = x_test_m)

# Create matrices
# Fit model (alpha=0.5 for Elastic Net)
sm_cv_model_osm <- cv.glmnet(x_train_matrix_m, sm_train, alpha = 0.5)

# Predict
sm_enet_pred_osm <- predict(sm_cv_model_osm, newx = x_test_matrix_m, s = "lambda.min")

# Evaluate performance with RMSE (Root Mean Squared Error)
sm_enet_rmse_osm <- sqrt(mean((sm_enet_pred_osm - sm_test)^2))
print(paste("RMSE for Elastic Net Regression:", sm_enet_rmse_osm))
```



```{r xgb}
library(xgboost)

# Train an XGBoost model
sm_xgb_model_osm <- xgboost(data = x_train_matrix_m, label = sm_train,
                         objective = "reg:squarederror", nrounds = 100, max.depth = 6, eta = 0.1)

# Predict
sm_xgb_pred_osm <- predict(sm_xgb_model_osm, newdata = x_test_matrix_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
sm_xgb_rmse_osm <- sqrt(mean((sm_xgb_pred_osm - sm_test)^2))
print(paste("RMSE for XGBoost Regression:", sm_xgb_rmse_osm))
```





## RELAPSE STATUS MODEL TRAINING

```{r LR}
rs_log_reg_model <- glm(rs_train ~ .,
                     data = x_train, family = "binomial")

# Summary of the model
summary(rs_log_reg_model)


# Predict on test data
predictions_status <- predict(rs_log_reg_model, newdata = x_test, type = "response")
pred_class_status <- ifelse(predictions_status > 0.5, 1, 0)  # 1 = Dead, 0 = Alive

# Confusion Matrix to evaluate the performance
rs_conf_matrix_status <- confusionMatrix(factor(pred_class_status), factor(rs_test))
print(rs_conf_matrix_status)

```


### Random Forest
```{r RF}
library(randomForest)

set.seed(123)
rs_rf_model <- randomForest(rs_train ~ .,
                         data = x_train, ntree = 500, importance = TRUE)

# Predict and evaluate
rs_rf_pred <- predict(rs_rf_model, newdata = x_test)

confusionMatrix(rs_rf_pred, factor(rs_test))

```



```{r GML}
library(glmnet)

# Fit model (alpha=0.5 for Elastic Net)
rs_cv_model <- cv.glmnet(x_train_matrix, rs_train, alpha = 0.5, family = "binomial")

# Predict
enet_pred <- predict(rs_cv_model, newx = x_test_matrix, s = "lambda.min", type = "response")
enet_pred_class <- ifelse(enet_pred > 0.5, 1, 0)
confusionMatrix(as.factor(enet_pred_class), as.factor(rs_test))
```



```{r xgg}

library(xgboost)


rs_xgb_model <- xgboost(data = x_train_matrix, label = as.numeric(rs_train) - 1,
                     objective = "binary:logistic", nrounds = 100, max.depth = 6, eta = 0.1)

# Prediction
xgb_pred <- predict(rs_xgb_model, newdata = x_test_matrix)
xgb_pred_class <- ifelse(xgb_pred > 0.5, 1, 0)
confusionMatrix(as.factor(xgb_pred_class), factor(rs_test))

```







## OVERALL RELAPSE  MONTHS
```{r LRR}
rm_linear_reg_model <- lm(rm_train ~ Age_at_Diagnosis + Tumor_Size + Cancer_Type_Detailed +
                         Lymph_nodes_examined_positive + 
                         Nottingham_prognostic_index + Tumor_Stage + Neoplasm_Histologic_Grade + ER + 
                         PR + HER2 + Chemo + Hormone + Radio + Menopause,
                       data = x_train_m)

# Summary of the model
summary(rm_linear_reg_model)

# Predict on test data
rm_predictions_osm <- predict(rm_linear_reg_model, newdata = x_train_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
rm_rmse_osm <- sqrt(mean((rm_predictions_osm - rm_test)^2))
print(paste("RMSE for Linear Regression:", rm_rmse_osm))

```



```{r}
library(randomForest)

# Train a Random Forest Regression model
rm_rf_reg_model <- randomForest(rm_train ~ Age_at_Diagnosis + Tumor_Size + Cancer_Type_Detailed + 
                             Lymph_nodes_examined_positive + Nottingham_prognostic_index + Tumor_Stage + 
                             Neoplasm_Histologic_Grade + ER + PR + HER2 + Chemo + Hormone + Radio + Menopause,
                             data = x_train_m, ntree = 500)

# Predict on test data
rm_rf_reg_pred <- predict(rm_rf_reg_model, newdata = x_train_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
rm_rf_reg_rmse <- sqrt(mean((rm_rf_reg_pred - rm_test)^2))
print(paste("RMSE for Random Forest Regression:", rm_rf_reg_rmse))
```

```{r}
library(glmnet)

# Create matrices
# Fit model (alpha=0.5 for Elastic Net)
rm_cv_model_osm <- cv.glmnet(x_train_matrix_m, rm_train, alpha = 0.5)

# Predict
rm_enet_pred_osm <- predict(rm_cv_model_osm, newx = x_train_matrix_m, s = "lambda.min")

# Evaluate performance with RMSE (Root Mean Squared Error)
rm_enet_rmse_osm <- sqrt(mean((rm_enet_pred_osm - rm_test)^2))
print(paste("RMSE for Elastic Net Regression:", rm_enet_rmse_osm))
```



```{r XGB}
library(xgboost)

# Train an XGBoost model
rm_xgb_model_osm <- xgboost(data = x_train_matrix_m, label = rm_train,
                         objective = "reg:squarederror", nrounds = 100, max.depth = 6, eta = 0.1)

# Predict
rm_xgb_pred_osm <- predict(rm_xgb_model_osm, newdata = x_train_matrix_m)

# Evaluate performance with RMSE (Root Mean Squared Error)
rm_xgb_rmse_osm <- sqrt(mean((rm_xgb_pred_osm - rm_test)^2))
print(paste("RMSE for XGBoost Regression:", rm_xgb_rmse_osm))
```







```{r classification_metric}
# Collect classification results
classification_results <- data.frame(
  Model = c(
    "Overall Survival - Logistic Regression",
    "Overall Survival - Random Forest",
    "Overall Survival - Elastic Net",
    "Overall Survival - XGBoost",
    "Relapse Status - Logistic Regression",
    "Relapse Status - Random Forest",
    "Relapse Status - Elastic Net",
    "Relapse Status - XGBoost"
  ),
  Accuracy = c(
    ss_conf_matrix_status$overall["Accuracy"],
    confusionMatrix(ss_rf_pred, factor(ss_test))$overall["Accuracy"],
    confusionMatrix(as.factor(ss_enet_pred_class), as.factor(ss_test))$overall["Accuracy"],
    confusionMatrix(as.factor(ss_xgb_pred_class), ss_test)$overall["Accuracy"],
    rs_conf_matrix_status$overall["Accuracy"],
    confusionMatrix(rs_rf_pred, factor(rs_test))$overall["Accuracy"],
    confusionMatrix(as.factor(enet_pred_class), as.factor(rs_test))$overall["Accuracy"],
    confusionMatrix(as.factor(xgb_pred_class), factor(rs_test))$overall["Accuracy"]
  ),
  Kappa = c(
    ss_conf_matrix_status$overall["Kappa"],
    confusionMatrix(ss_rf_pred, factor(ss_test))$overall["Kappa"],
    confusionMatrix(as.factor(enet_pred_class), as.factor(ss_test))$overall["Kappa"],
    confusionMatrix(as.factor(xgb_pred_class), ss_test)$overall["Kappa"],
    rs_conf_matrix_status$overall["Kappa"],
    confusionMatrix(rs_rf_pred, factor(rs_test))$overall["Kappa"],
    confusionMatrix(as.factor(enet_pred_class), as.factor(rs_test))$overall["Kappa"],
    confusionMatrix(as.factor(xgb_pred_class), factor(rs_test))$overall["Kappa"]
  )
)

# Print the table
print(classification_results)

```


```{r}
# Load necessary library
library(ggplot2)
library(tidyr)

# Convert the classification_results data frame into a long format for easier plotting
classification_results_long <- classification_results %>%
  gather(key = "Metric", value = "Value", Accuracy, Kappa)

# Plotting comparison of Accuracy and Kappa for each model
ggplot(classification_results_long, aes(x = Model, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Model Comparison: Accuracy and Kappa",
       x = "Model", y = "Value") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.title = element_text(size = 12)) +
  scale_fill_manual(values = c("Accuracy" = "blue", "Kappa" = "orange")) +
  geom_text(aes(label = round(Value, 2)), position = position_dodge(0.8), vjust = -0.3, size = 4)

```

```{r}
# Collect regression results
regression_results <- data.frame(
  Model = c(
    "Overall Survival Months - Linear Regression",
    "Overall Survival Months - Random Forest",
    "Overall Survival Months - Elastic Net",
    "Overall Survival Months - XGBoost",
    "Relapse Months - Linear Regression",
    "Relapse Months - Random Forest",
    "Relapse Months - Elastic Net",
    "Relapse Months - XGBoost"
  ),
  RMSE = c(
    sm_rmse_osm,
    sm_rf_reg_rmse,
    sm_enet_rmse_osm,
    sm_xgb_rmse_osm,
    rm_rmse_osm,
    rm_rf_reg_rmse,
    rm_enet_rmse_osm,
    rm_xgb_rmse_osm
  )
)

# Print the table
print(regression_results)

```

```{r}

# Load necessary library
library(ggplot2)

# Plotting RMSE comparison for each model
ggplot(regression_results, aes(x = Model, y = RMSE, fill = Model)) +
  geom_bar(stat = "identity") +
  labs(title = "Model Comparison: RMSE for Regression Models",
       x = "Model", y = "RMSE") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.title = element_text(size = 12)) +
  scale_fill_viridis_d() +  # Optional color scale for better aesthetics
  geom_text(aes(label = round(RMSE, 2)), vjust = -0.3, size = 4)
```

## SAVING THE TOP MODEL
```{r}
saveRDS(ss_rf_model, "ss_rf_model.rds")
saveRDS(ss_xgb_model, "ss_xgb_model.rds")
saveRDS(sm_rf_reg_model, "sm_rf_reg_model.rds")
saveRDS(rs_cv_model, "rs_cv_model.rds")
saveRDS(rs_xgb_model, "rs_xgb_model.rds")
saveRDS(rm_rf_reg_model, "rm_rf_reg_model.rds")
```




# SIMULATION

```{r new_data}
# Simulate a new data point for prediction
new_data <- data.frame(
  Age_at_Diagnosis = 50,
  Tumor_Size = 3.5,  # Example size in cm
  Cancer_Type_Detailed = 3,  # E.g., "Breast Invasive Ductal Carcinoma"
  Lymph_nodes_examined_positive = 2,
  Nottingham_prognostic_index = 4.5,
  Tumor_Stage = 2,  # Tumor stage as a factor
  Neoplasm_Histologic_Grade = 2,  # Example grade
  ER = 1,  # ER positive
  PR = 1,  # PR positive
  HER2 = 0,  # HER2 negative
  Chemo = 1,  # Chemotherapy received
  Hormone = 0,  # Hormone therapy not received
  Radio = 1,  # Radiotherapy received
  Menopause = 1  # Post-menopausal
)



# View the new data
new_data
```


```{r}

# Convert Tumor Stage to a factor with appropriate labels
new_data$Tumor_Stage <- factor(new_data$Tumor_Stage,
                                levels = c(1, 2, 3, 4),
                                labels = c("Stage I", "Stage II", "Stage III", "Stage IV"))


# Encode ER/PR/HER2 status as binary (Positive = 1, Negative = 0)
new_data$ER <- factor(new_data$ER, levels = c(0, 1))
new_data$PR <- factor(new_data$PR,levels = c(0, 1))
new_data$HER2 <- factor(new_data$HER2, levels = c(0, 1))

new_data$Neoplasm_Histologic_Grade <- factor(new_data$Neoplasm_Histologic_Grade, levels = c(1:3))

# Encode Menopausal state as binary
new_data$Menopause <- factor(new_data$Menopause, levels = c(0, 1))

# Encode therapies (Chemotherapy, Hormone Therapy, Radiotherapy) as binary
new_data$Chemo <- factor(new_data$Chemo, levels = c(0, 1))
new_data$Hormone <- factor(new_data$Hormone, levels = c(0, 1))
new_data$Radio <- factor(new_data$Radio, levels = c(0, 1))

# Convert Cancer Type Detailed to factor and numeric
new_data$Cancer_Type_Detailed <- factor(new_data$Cancer_Type_Detailed, levels = c(1:7),
                                         labels = c("Breast", "Breast Invasive Ductal Carcinoma",
                                                    "Breast Invasive Lobular Carcinoma", "Breast Invasive Mixed Mucinous Carcinoma",
                                                    "Breast Mixed Ductal and Lobular Carcinoma", "Invasive Breast Carcinoma", "Others"))




# Ensure necessary columns are present
new_data <- new_data %>%
  select(Age_at_Diagnosis, Tumor_Size, Cancer_Type_Detailed,
         Lymph_nodes_examined_positive, Nottingham_prognostic_index, Tumor_Stage, 
         Neoplasm_Histologic_Grade, ER, PR, HER2, Chemo, Hormone, Radio, Menopause)
head(new_data)
```







### Relapse Free Status
```{r}
# Convert new_data to a numeric matrix
new_x <- model.matrix(~ . - 1, data = new_data)

# Now predict
rs_prediction <- predict(rs_cv_model, newx = new_x, s = "lambda.min", type = "response")

# Output the predicted survival months
rs_prediction
```


### Relapse Free Months
```{r}

# Predict survival status using the pre-trained Elastic Net model (cv_model)
rm_prediction <- predict(rm_rf_reg_model, newdata = new_data)


# Output the predicted survival status
rm_prediction
```








### Survival Months
```{r}

#setequal(colnames(x_new_data), colnames(x_train_matrix))  # Should return TRUE

# Predict survival months using the pre-trained XGBoost model (xgb_model_osm)
sm_prediction <- predict(sm_rf_reg_model, newdata = new_data)

# Output the predicted survival months
sm_prediction
```


### Survival  Status
```{r}
# Create matrix for Elastic Net prediction of Survival Status
x_new_data <- model.matrix(~ . -1, data = new_data)

# Rename new data columns to match training data
colnames(x_new_data)[colnames(x_new_data) == "ERYes"] <- "ER1"
colnames(x_new_data)[colnames(x_new_data) == "PRYes"] <- "PR1"
colnames(x_new_data)[colnames(x_new_data) == "HER2Yes"] <- "HER21"
colnames(x_new_data)[colnames(x_new_data) == "ChemoYes"] <- "Chemo1"
colnames(x_new_data)[colnames(x_new_data) == "HormoneYes"] <- "Hormone1"
colnames(x_new_data)[colnames(x_new_data) == "RadioYes"] <- "Radio1"
colnames(x_new_data)[colnames(x_new_data) == "MenopausePost"] <- "Menopause1"


# Predict survival status using the pre-trained Elastic Net model (cv_model)
ss_prediction <- predict(ss_rf_model, newdata = new_data)


# Output the predicted survival status
ss_prediction
```





